---
output:
  pdf_document: default
  html_document: default
---
## Assignment 9
### Daniel DeBonis

### Your task is to choose one of the New York Times APIs, construct an interface in R to read in the JSON data, and transform it into an R DataFrame.

Certian packages are necessary in dealing with the JSON format of the API connection.
```{r}
library(jsonlite)
library(httr)
library(httr2)
```

I selected the Top Stories API. I decided to search for the top stories that are considered World News. 
```{r}
worldraw <- GET("https://api.nytimes.com/svc/topstories/v2/world.json?api-key=V3mpHG49JtsZy4eL53sGtlphnbDHZab3")
worldtext <- content(worldraw, as = 'text') 
worldjson <- fromJSON(worldtext)
worlddf <- worldjson$results
```

```{r}
head(worlddf)
```

The linked API had a lot of information on each article, such as the region of the world concerned, title, author(s), links, and many other variables of interest. If I wanted to do further analysis, an important step would be to specify only the relevant columns be included upon converting from JSON to a data frame in R.